---
title: "Lab 08 - Text Mining/NLP"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = T, include  = T)
```

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and ngrams from text
- Use dplyr and ggplot2 to analyze and visualize text data
- Try a theme model using `topicmodels`

# Lab description

For this lab we will be working with the medical record transcriptions from https://www.mtsamples.com/ available at https://github.com/JSC370/JSC370-2025/tree/main/data/medical_transcriptions.

# Deliverables

1. Questions 1-7 answered, knit to pdf or html output uploaded to Quercus.

2. Render the Rmarkdown document using `github_document` and add it to your github site. Add link to github site in your html.


### Setup packages

You should load in `tidyverse`, (or `data.table`), `tidytext`, `wordcloud2`, `tm`, and `topicmodels`.


## Read in the Medical Transcriptions

Loading in reference transcription samples from https://www.mtsamples.com/

```{r}
library(tidytext)
library(tidyverse)
library(wordcloud2)
library(tm)
library(topicmodels)

mt_samples <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/medical_transcriptions/mtsamples.csv")
mt_samples <- mt_samples |>
  select(description, medical_specialty, transcription)

head(mt_samples)
```

---

## Question 1: What specialties do we have?

We can use `count()` from `dplyr` to figure out how many different medical specialties are in the data. Are these categories related? overlapping? evenly distributed? Make a bar plot.

```{r}
mt_samples |>
  count(medical_specialty, sort = TRUE) |>
  ggplot(aes(fct_reorder(medical_specialty, n), n)) + 
  geom_col(fill = "dodgerblue") + 
  coord_flip() +  # Flip coordinates for better readability
  labs(title = "Distribution of Medical Specialties", 
       x = "Medical Specialty", 
       y = "Count") + 
  theme_bw()
```
```{r}
library(dplyr)

# Count occurrences of each medical specialty and sort by frequency
medical_specialty_counts <- mt_samples |> 
  count(medical_specialty, sort = TRUE)

# Print the counts
print(medical_specialty_counts)
```
There are 30 different medical specialties are in the data. Categories are not related and from the bar plot we can observe the Surgery is dominating.

---

## Question 2: Tokenize

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words with a bar plot
- Create a word cloud of the top 20 most frequent words

### Explain what we see from this result. Does it makes sense? What insights (if any) do we get?
The top words are mostly junk words and not informative, we would like to filter out those junk words.

```{r}
tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  group_by(word) |>
  summarize(word_frequency = n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)
tokens
```

```{r}
tokens |>
  ggplot(aes(fct_reorder(word, word_frequency),word_frequency)) +
  geom_bar(stat = "identity", fill ="dodgerblue") +
  coord_flip()+
  theme_bw()
```
```{r}
tokens |>
  count(word, sort = TRUE) |>
  wordcloud2(size = 0.3, color = "random-light", backgroundColor = "dodgerblue")
```
---

## Question 3: Stopwords

- Redo Question 2 but remove stopwords
- Check `stopwords()` library and `stop_words` in `tidytext`
- Use regex to remove numbers as well
- Try customizing your stopwords list to include 3-4 additional words that do not appear informative

### What do we see when you remove stopwords and then when you filter further? Does it give us a better idea of what the text is about?
Words become more meaningful, yes.

```{r}
head(stopwords("english"))
length(stopwords("english"))
head(stop_words)

tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription, token = "words") |>
  anti_join(stop_words, by = "word") |>  # Remove stopwords
  filter(!str_detect(word, "^[0-9]+$")) |>  # Remove numbers
  filter(!word %in% c("mm", "mg", "noted")) |>
  count(word, sort = TRUE) |>  # Count occurrences
  top_n(20, n)  # Select top 20 most frequent words

# Bar Plot of Top 20 Most Frequent Words
tokens |>
  ggplot(aes(x = n, y = fct_reorder(word, n))) +
  geom_col(fill = "dodgerblue") +
  labs(title = "Top 20 Most Frequent Words (After Stopword Removal)",
       x = "Frequency",
       y = "Word") +
  theme_bw()

tokens |>
  count(word, sort = TRUE) |>
  wordcloud2(size = 0.4, color = "random-light", backgroundColor = "dodgerblue")
```

---

## Question 4: ngrams

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams? Note we need to remove stopwords a little differently. You don't need to recreate the wordclouds.

```{r}
stop_words2 <- c(stop_words$word, "mm", "mg", "noted")
sw_start <- paste0("^", paste(stop_words2, collapse=" |^"), "$")
sw_end <- paste0("", paste(stop_words2, collapse="$| "), "$")

tokens_bigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 2) |>
  filter(!grepl(sw_start, ngram, ignore.case = TRUE)) |> 
  filter(!grepl(sw_end, ngram, ignore.case = TRUE)) |> 
  filter(!grepl("[[:digit:]]+", ngram)) |> 
  group_by(ngram) |>
  summarize(word_frequency = n()) |> 
  arrange(across(word_frequency, desc)) |>
  head(20)

# Bar Plot of Top 20 Bi-Grams
tokens_bigram |>
  ggplot(aes(ngram, word_frequency)) +
  geom_col(fill = "dodgerblue") +
  coord_flip()+
  theme_bw()
```
---

## Question 5: Examining words

Using the results from the bigram, pick a word and count the words that appear before and after it, and create a plot of the top 20.

```{r}
library(stringr)
tokens_bigram |>
  filter(str_detect(ngram, regex("\\sblood$|^blood\\s"))) |>
    mutate(word = str_remove(ngram, "blood"),
         word = str_remove_all(word, " ")) |>
  group_by(word) |>
  head(20)|>
  ggplot(aes(reorder(word, word_frequency), word_frequency)) +
  geom_col(fill = "dodgerblue")+
  coord_flip()+
  theme_bw()
```
---

## Question 6: Words by Specialties

Which words are most used in each of the specialties? You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stopwords. How about the 5 most used words?
Results are shown below

```{r}
mt_samples |>
   unnest_tokens(word, transcription) |>
   anti_join(stop_words, by = "word") |>  # Remove stopwords
   filter(!str_detect(word, "^[0-9]+$")) |>  # Remove numbers
   filter(!word %in% c("mm", "mg", "noted")) |>  # Remove custom stopwords
   count(medical_specialty, word, sort = TRUE) |>  # Count words per specialty
   group_by(medical_specialty) |> 
   slice_max(order_by = n, n = 5) |>  # Select top 5 words per specialty
   ungroup()
```


## Question 7: Topic Models

See if there are any themes in the data by using a topic model (LDA). 

- you first need to create a document term matrix
- then you can try the LDA function in `topicmodels`. Try different k values. k is the number of clusters.
- create a facet plot of the results from the LDA (see code from lecture)


```{r}
transcripts_dtm <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  anti_join(stop_words, by = "word") |>
  filter(!str_detect(word, "^[0-9]+$")) |>
  filter(!word %in% c("mm", "mg", "noted")) |>
  DocumentTermMatrix()
```

```{r}
transcripts_dtm <- as.matrix(transcripts_dtm)   

transcripts_lda <- LDA(transcripts_dtm, k = 3, control = list(seed=1234))

transcripts_top_terms <- tidy(transcripts_lda, matrix = "beta") |>
  filter(!str_detect(term, "^[0-9]+$")) |>
  group_by(topic) |>
  slice_max(beta, n=10)|>
  ungroup()|>
  arrange(topic, -beta)

transcripts_top_terms |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE)+
  facet_wrap(~topic, scales="free")+
  scale_y_reordered()+
  theme_bw()
```




