---
title: "Lab 9 - HPC"
output: 
html_document: default
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

# Learning goals

In this lab, you are expected to practice the following skills:

-   Evaluate whether a problem can be parallelized or not.
-   Practice with the parallel package.
-   Use Rscript to submit jobs.

```{r echo=FALSE, eval = FALSE}
# install any missing packages
install.packages("microbenchmark")
```

## Problem 1

Give yourself a few minutes to think about what you learned about
parallelization. List three examples of problems that you believe may be
solved using parallel computing, and check for packages on the HPC CRAN
task view that may be related to it.

1.  **Monte Carlo Simulations**
    -   Monte Carlo methods are computationally intensive and often
        involve running a large number of independent simulations.
        Parallelizing these simulations can significantly speed up the
        computation.\
    -   **Related HPC CRAN Packages:** `parallel`, `foreach`,
        `doParallel`, `RcppParallel`
2.  **Large-Scale Machine Learning Training**
    -   Training deep learning models or running ensemble methods (e.g.,
        Random Forest, XGBoost) on large datasets benefits from parallel
        computation to optimize resource usage and reduce training
        time.\
    -   **Related HPC CRAN Packages:** `caret`, `xgboost`,
        `randomForest`, `h2o`
3.  **Genomic Data Processing and Bioinformatics Analysis**
    -   Analyzing large-scale sequencing data (e.g., DNA sequencing,
        RNA-Seq) requires parallelization to handle massive datasets
        efficiently, such as genome-wide association studies (GWAS).\
    -   **Related HPC CRAN Packages:** `BiocParallel`, `future`, `snow`,
        `BatchJobs`

## Problem 2: Pre-parallelization

The following functions can be written to be more efficient without
using `parallel`:

1.  This function generates a `n x k` dataset with all its entries
    having a Poisson distribution with mean `lambda`.

```{r p2-fun1}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  # YOUR CODE HERE
  matrix(rpois(n*k, lambda = lambda), ncol = k)
  # allow us to pre-allocating memory
}

# Benchmarking
bench <- microbenchmark::microbenchmark(
  fun1(100),
  fun1alt(100),
  unit = "ns"
)
```

How much faster?

To determine how much faster `fun1alt(100)` is compared to `fun1(100)`,
we calculate the speedup factor:

$$
\text{Speedup} = \frac{\text{Mean execution time of fun1}}{\text{Mean execution time of fun1alt}}
$$

From the benchmarking results: - Mean execution time of `fun1(100)`:
**202,365.34 ns** - Mean execution time of `fun1alt(100)`: **23,648.39
ns**

$$
\text{Speedup} = \frac{202365.34}{23648.39} \approx 8.56
$$

Thus, `fun1alt(100)` is approximately **8.56 times faster** than
`fun1(100)`. The improvement comes from pre-allocating memory using
`matrix()` instead of dynamically growing an object with `rbind()`,
which is inefficient.

2.  Find the column max (hint: Checkout the function `max.col()`).

```{r p2-fun2}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  # YOUR CODE HERE
  x[cbind(max.col(t(x)), 1:ncol(x))]
  # avoid function calls inside loops
}

# Benchmarking
bench <- microbenchmark::microbenchmark(
  fun2(x),
  fun2alt(x),
  unit = "us"
)
```

*Answer here with a plot.*
```{r}
plot(bench)
ggplot2::autoplot(bench) + ggplot2::theme_minimal()
```

## Problem 3: Parallelize everything

We will now turn our attention to non-parametric
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)).
Among its many uses, non-parametric bootstrapping allow us to obtain
confidence intervals for parameter estimates without relying on
parametric assumptions.

The main assumption is that we can approximate many experiments by
resampling observations from our original dataset, which reflects the
population.

This function implements the non-parametric bootstrap:

```{r p3-boot-fun}
library(parallel)
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
 
  # Making the cluster using `ncpus`
  # STEP 1: GOES HERE
  cl <- makePSOCKcluster(ncpus)
  # create workers
  # creating cluster for parallel computing
  # ncpus specifying using multiple CPU cores
  # PSOCK parallel socket cluster
  # STEP 2: GOES HERE
  # on.exit(stopCluster(cl))
  # export the variable to the cluster
  clusterExport(cl, varlist = c("idx", "dat", "stat"), envir = environment())
  # sending the variables to all workers nodes
  # each run in isolated environment, don't have access to global variable
  # idx -> resampling indices for bootstrapping
  # dat -> dataset
  # stat -> the statistical function that we use to compute the estimates

  
  # STEP 3: THIS FUNCTION NEEDS TO BE REPLACED WITH parLapply
  ans <- lapply(seq_len(R), function(i) {
    stat(dat[idx[,i], , drop=FALSE])
  })
  
  # Coercing the list into a matrix
  ans <- do.call(rbind, ans)
  
  # STEP 4: GOES HERE
  stopCluster(cl)
  # why?
  # free up system resources
  ans
  
}
```

1.  Use the previous pseudocode, and make it work with `parallel`. Here
    is just an example for you to try:

```{r p3-test-boot}
# Bootstrap of a linear regression model
my_stat <- function(d) coef(lm(y~x, data = d)) 

# DATA SIM
set.seed(1)
n <- 500 
R <- 1e4
x <- cbind(rnorm(n)) 
y <- x*5 + rnorm(n)

# Check if we get something similar as lm
ans0 <- confint(lm(y~x))
cat("OLS CI \n")
ans0

ans1 <- my_boot(dat = data.frame(x,y), my_stat, R=R, ncpus = 4)
qs <- c(0.025, 0.975)
cat("Bootstrap CI \n")
print(t(apply(ans1, 2, quantile, probs = qs)))
```

2.  Check whether your version actually goes faster than the
    non-parallel version:

```{r benchmark-problem3}
parallel::detectCores()
system.time(my_boot(dat = data.frame(x, y), my_stat, R=4000, ncpus = 1L))
system.time(my_boot(dat = data.frame(x, y), my_stat, R=4000, ncpus = 8L))
```

Since the 8 cores is running faster, so the first line is actually the result running on 8 cores. So it has smaller values for 8 cores comparing to only 1 core.

## Problem 4: Compile this markdown document using Rscript

Once you have saved this Rmd file, try running the following command in
your terminal:

``` bash
Rscript --vanilla -e 'rmarkdown::render("lab09-hpc.Rmd")' &
```

Where `[full-path-to-your-Rmd-file.Rmd]` should be replace with the full
path to your Rmd file... :).
